Introduction:

Dark data is defined by @craiu2023 as "any kind of data that are lost or distorted before analysis". The most common forms of dark data that data scientists and statisticians come across include missing data, as well as censored data, both of which will be the main discussion points in this guide. By the end of this user guide, you should be able to identify these two types of dark data, and also have an increased set of skills to be able to handle them. Check out

Pre-Requisites: In order to be able to replicate and follow along with the steps and functions that will be discussed in this user guide, it is assumed that the user has RStudio downloaded, as well as the latest version of R installed. Also, it will be assumed that you have a basic understanding of common commands in R, such as installing packages and basic data handling techniques, as well as data visualization (ggplot2).

Steps(with code and graphics examples):

-Naniar package (missing data)

The naniar package is designed specifically to help visualize missing data. More specifically, it assists in helping deal with NA values that show up in certain data sets. These missing values often show up in certain circumstances such as questionnaires, where the respondent chooses not to answer, as well as things such as technological issues and participation dropouts. In these steps, the main goal is to show you how to make simple visual plots that make it easy to identify and locate missing values in a data set. @visdat has written an article that explains many different ways that the naniar package can be utilized in the data science world.

1.  Install and load necessary packages and read in your data (I'll be using the "airquality" data set in the naniar package for this demonstration). This data set is built into the naniar package and it observes 153 measurements of air qaulity in NYC over the summer of 1973.

```{r}
# install.packages(naniar)
# install.packages(ggplot2)
library(naniar)
library(ggplot2)
data <- airquality
```

2.  Create a scatter plot using geom_point in the ggplot package displaying the relationship between the Solar.R variable (assign to x-axis) and the Ozone variable (y-axis).

```{r}
ggplot(data, aes(x=Solar.R, y=Ozone)) +
         geom_point() +
  ggtitle("Scatterplot of Solar vs Ozone")
  
```

This graph will exclude NA values and instead will tell you that 42 rows have been removed because they contain missing values. This doesn't give you anything to work with regarding further analysis of these missing values.

3.  Use gg_miss_var to visualize the missing data.

```{r}
gg_miss_var(airquality)
```

This function counts the missing variables that are found in each variable of the data set and gives you a simple chart to visualize where these missing values are located. This visualization can come in handy for experimental designing and determining which aspects of the experiment gather the most missing values and adjustments can be made accordingly.

4.  Create a data frame using miss_var_summary

```{r}
miss_var_summary(airquality)
```

This table also displays the amount of missing values in each variable, except now we've added a new column that tells us the percentage of all the values in each variable that are missing. When thinking about handling missing data, it is important to consider the context of why this data is important. A table will make it easier to display information to other people, but a graph may be more helpful to help locate where these missing values are at.

-Survival package The survival package is designed to help statisticians deal with censored data. Censored data occurs when data is incomplete or partially known. This will most commonly happen when dealing with topics such as machine failure and death. The three main components of censored data includes left censoring (the event occurred before the study period), right censoring (the event occurred after the study period), and interval censoring (event occurred between two certain points, but that point is unknown). For a more in depth breakdown on the function of this package, @rich2010 has made an article that takes you further in depth into overall survival analysis and other ways to view/handle censored data.

1.  Install and load the necessary packages (the data set in the survival package will be used in this guide). The lung data set in the survival package examines 228 advanced lung cancer patients and their survival times. Form this data, we will examine censored data, and in this case, censored data means that the patient died before the study was concluded.

```{r}
# install.packages(survival)
# install.packages(dplyr)
# install.packages("ggsurvfit")
library(survival)
library(dplyr)
library(ggsurvfit)
data2 <- lung
```

2.  Use the "Surv" function to look at survival times of the first 10 patients.

```{r}
Surv(lung$time, lung$status)[1:10]
```

In this table, a plus sign next to a value will indicate that the value is right censored (patient died after the study period).

3.  Knowing which values are censored in a data set can help with further analysis. Now, we are going to look at the amount of time in days until an event (in this case, a patient dying) by creating a Kaplan-Meier plot. The purpose of this plot is to show us the survival probability of patients as time goes on based on sex. To do this, we will use the "survfit2" function to create a survival curve and assign our two variables to the graph. Also, we will use "add_confidence_interval" and "add_risk_table" to give us a better understanding of how the data can fluctuate and how many values we are seeing at each time stamp.

```{r}
survfit2(Surv(time, sex) ~ 1, data = lung) |> 
  ggsurvfit() +
  labs(x = "Days", y = "Overall survival probability") + 
  add_confidence_interval() +
  add_risktable()
```

Conclusion After reading this user guide, you should hopefully have a better idea on how to identify these two types of dark data, as well as knowing different ways of handling and analyzing them. To many, these data points may seem irrelevant or unimportant, but those people don't see how these values can play a crucial role in the overall results of a data set, as well as the impact they could have on future experiments and studies. The naniar package will come in handy for a variety of applications, as missing values are found a lot more frequently then censored data. However, that does not mean that knowing how to handle censored data will not be useful, as in this case where we are in the healthcare field, it will also play a crucial role in other areas of our lives. IF we can increase our awareness and knowledge of the many types of dark data, it will make our jobs as data scientists much easier and it will likely give us better and more accurate results for our studies.
